Design and Implementation of Probabilistic Programming Language Anglican

### 2017-02-25 paper/paper.tex
edits to section 9

Macro \texttt{defdist} should be used to define distributions.
\texttt{defdist} takes care of defining a separate type
for every distribution so that Clojure multimethods (or overloaded
methods) can be dispatched on
distribution types when needed, e.g. for custom proposal distributions 
used in an inference algorithm. The
Bernoulli distribution could be defined as follows:
\begin{lstlisting}[style=default]
(defdist bernoulli
  "Bernoulli distribution"
  [p]
  (sample* [this] (if (< (rand) p) 1 0))
  (observe* [this value]
    (Math/log (case value
                1 p
                0 (- 1. p)
                0.))))
\end{lstlisting}

In addition to distributions, Anglican provides \textit{random processes},
which define sequences of random variables that are not independent and
identically distributed. Random processes are similar to the so called
`exchangeable random primitives' in Church \cite{GMR+08} and
Venture \cite{MSP14}. However, random sequences generated by Anglican random 
processes are not required to be exchangeable. Random processes are
defined using the \texttt{defproc} macro and implement the
\texttt{anglican.runtime/{\linebreak[0]}random-process} protocol. This
protocol has two methods
\begin{itemize}
\item \texttt{produce}, which returns the distribution on the next
random variable in the sequence, and
\item \texttt{absorb}, which incorporates the value for the next random
variable and returns an updated random process.
\end{itemize}
 As an example, here is a definition of a beta-Bernoulli process, in which
each random variable is distributed according to a Bernoulli distribution with
an unknown parameter that is drawn from a beta distribution:
\begin{lstlisting}[style=default]
(defproc beta-bernoulli
  "beta-Bernoulli process"
  [a b]
  (produce [this] (bernoulli (/ a (+ a b))))
  (absorb [this x]
    (case x
      0 (beta-bernoulli a (inc b))
      1 (beta-bernoulli (inc a) b))))
\end{lstlisting}
Unlike typical implementations of exchangeable random processes,
Anglican's random processes do not have mutable state. The
\texttt{produce} and \texttt{absorb} methods are deterministic
and functional, and therefore do not have corresponding special
forms in Anglican. A sequence of random values can be generated
using a recursive loop in which \texttt{absorb} returns the
updated process for the next iteration. For example:
\begin{lstlisting}[style=default]
(defm sample-beta-binomial [n a b]
  (loop [process (beta-bernoulli a b)
         values []]
    (if (= (count values) n)
      values
      (let [dist (produce process)
            value (sample dist)]
        (recur (absorb process value)
               (conj values value))))))
\end{lstlisting}
Similarly, random processes can also be used to recursively
observe a sequence of values:
\begin{lstlisting}[style=default]
(loop [process (beta-bernoulli 1.0 1.0)
       [value0 & values0] values] 
  (if (= (count values0) 0)
    nil
    (let [dist (produce process)]
      (observe dist value0)
      (recur (absorb process value0) values0))))
\end{lstlisting}

@@

We run this program for \texttt{n} equal to 10, 15, 20, and 25. Clojure
is consistently only twice as fast as Anglican
(Table~\ref{tbl:hanoi}); considering the amount of overhead
Anglican introduces, this is a good result.
\begin{table}
    \centering
    \begin{tabular}{r|r|r|r|r}
        {\bf n}  & 10 & 15 & 20 & 25 \\ \hline
        {\bf Anglican} & 0.63 & 3.00 & 75.9 & 2381.  \\ \hline
        {\bf Clojure} &  0.32 & 1.45 & 38.1  &  1245.
    \end{tabular}
    \caption{Towers of Hanoi. Running times, in
    seconds, of Clojure and Anglican, averaged over 100 runs.
    Clojure is only twice as fast as Anglican.}
    \label{tbl:hanoi}
\end{table}

>>>>

The macro \texttt{defdist} is used to define distributions.
\texttt{defdist} takes care of defining a separate type
for every distribution so that Clojure multimethods (or overloaded
methods) can be dispatched on
distribution types when needed, e.g. for custom proposal distributions 
used in an inference algorithm. The
Bernoulli distribution could be defined as follows:
\begin{lstlisting}[style=default]
(defdist bernoulli
  "Bernoulli distribution"
  [p]
  (sample* [this] (if (< (rand) p) 1 0))
  (observe* [this value]
    (Math/log (case value
                1 p
                0 (- 1. p)
                0.))))
\end{lstlisting}

In addition to distributions, Anglican provides \textit{random processes},
which define sequences of random variables that are not independent and
identically distributed. Random processes are similar to the so called
`exchangeable random primitives' in Church \cite{GMR+08} and
Venture \cite{MSP14}. However, random sequences generated by Anglican random 
processes are not required to be exchangeable. Random processes are
defined using the \texttt{defproc} macro and implement the
\texttt{anglican.runtime/{\linebreak[0]}random-process} protocol. This
protocol has two methods
\begin{itemize}
\item \texttt{produce}, which returns the distribution on the next
random variable in the sequence, and
\item \texttt{absorb}, which incorporates the value for the next random
variable and returns an updated random process.
\end{itemize}
 As an example, here is a definition of a beta-Bernoulli process, in which
each random variable is distributed according to a Bernoulli distribution with
an unknown parameter that is drawn from a beta distribution:
\begin{lstlisting}[style=default]
(defproc beta-bernoulli
  "beta-Bernoulli process"
  [a b]
  (produce [this] (bernoulli (/ a (+ a b))))
  (absorb [this x]
    (case x
      0 (beta-bernoulli a (inc b))
      1 (beta-bernoulli (inc a) b))))
\end{lstlisting}
Unlike typical implementations of exchangeable random processes,
Anglican's random processes do not have mutable state. The
\texttt{produce} and \texttt{absorb} methods are deterministic
and functional, and therefore do not have corresponding special
forms in Anglican. A sequence of random values can be generated
using a recursive loop in which \texttt{absorb} returns the
updated process for the next iteration. For example:
\begin{lstlisting}[style=default]
(defm sample-beta-binomial [n a b]
  (loop [process (beta-bernoulli a b)
         values []]
    (if (= (count values) n)
      values
      (let [dist (produce process)
            value (sample dist)]
        (recur (absorb process value)
               (conj values value))))))
\end{lstlisting}
Similarly, random processes can also be used to recursively
observe a sequence of values:
\begin{lstlisting}[style=default]
(defm observe-beta-binomial [values a b]
  (loop [process (beta-bernoulli a b)
         values values] 
    (when (not-empty values)
      (let [dist (produce process)
            value (first values)]
        (observe dist value)
        (recur (absorb process value)
               (rest values))))))
\end{lstlisting}

@@

We run this program for \texttt{n} equal to 10, 15, 20, and 25. Clojure
is consistently only twice as fast as Anglican
(Table~\ref{tbl:hanoi}); considering the amount of overhead
Anglican introduces, this is a good result.
\begin{table}[!b]
    \centering
    \begin{tabular}{r|r|r|r|r}
        {\bf n}  & 10 & 15 & 20 & 25 \\ \hline
        {\bf Anglican} & 0.63 & 3.00 & 75.9 & 2381.  \\ \hline
        {\bf Clojure} &  0.32 & 1.45 & 38.1  &  1245.
    \end{tabular}
    \caption{Towers of Hanoi. Running times, in
    seconds, of Clojure and Anglican, averaged over 100 runs.
    Clojure is only twice as fast as Anglican.}
    \label{tbl:hanoi}
\end{table}
